Polynomial Time and Dependent Types

### 2023-10-25 paper/paper.tex
[ paper ] I was right
[ paper ] Fix acks

We will see in \autoref{sec:consfree-sound} that this system is sound
for polytime by a realisability argument. Completeness can be seen
more directly by constructing a function that iterates a function for
a statically known polynomial number of times in the size of the
input. Assume that we have a known polynomial
$p(n) = c_dn^d + \dots + c_0$ of degree $d$ with natural number
coefficients and some single step function
$f : \mathit{St} \multimap \mathit{St}$ over a state type
$\mathit{St}$ that runs to completion for input of size $n$ in $p(n)$
steps. Then, using the iterator above we can iterate $f$ over a
$\tyNat$ representing the size of the input:\bob{CHECK THIS}
\begin{displaymath}
  \begin{array}{l}
    I_1 : \tyNat \multimap \mathit{St} \multimap \mathit{St} \\
    I_1 = \lambda n. \lambda s.\tmRec\,n\,\{\conZero \mapsto s; \conSucc(s) \mapsto f\,s \}
  \end{array}
\end{displaymath}
To achieve higher degrees, we can use $\dupNat$ to nest iterations:
\begin{displaymath}
  \begin{array}{l}
    I_{k+1} : \tyNat \multimap \mathit{St} \multimap \mathit{St} \\
    I_{k+1} = \lambda n. \lambda s.
    \begin{array}[t]{@{}l}
      \Let\,(n,n') = \dupNat\,n\,\In\\
      \tmRec\,n\,\{\conZero \mapsto s; \conSucc(s) \mapsto I_k\,n'\,s \}
    \end{array}
  \end{array}
\end{displaymath}
By further use of $\dupNat$ and composition to handle addition of
polynomials, the function $f$ can now be iterated $p(n)$ many times,
where $n$ is the input $\tyNat$. Thus, the \emph{Cons-free} system can
represent all polytime computations.

@@

  \inferrule*
  {\Gamma_1 \vdash M : \Diamond \\ \Gamma_2 \vdash N : \tyNat}
  {\Gamma_1, \Gamma_2 \vdash \conSucc(M, N) : \tyNat}
\end{mathpar}
To construct a $\conZero$, we must have a $\Diamond$ to pay for it,
and likewise, to construct a $\conSucc$ we must pay a $\Diamond$. We
can think of $\Diamond$s as an unit of iterable data. Iterability is
``saved up'' in data during construction, and released during
iteration. Diamonds cannot be created by a program itself, for the
same reason that constructors were prohibited in the \emph{Cons-free}
system, but they are released from iterable data during iteration. The
LFPL natural number iterator has the following typing rule:
\begin{displaymath}
  \inferrule*
  {d : \Diamond \vdash M_z : A \\ d : \Diamond, x : A \vdash M_s : A \\ \Gamma \vdash N : \tyNat}
  {\Gamma \vdash \tmRec\,N\,\{\conZero(d) \mapsto M_z; \conSucc(d,x) \mapsto M_s \} : A}
\end{displaymath}
The difference with the \emph{Cons-free} iterator above is that the
$\conZero$ and $\conSucc$ cases now both have an additional binding of
type $\Diamond$. This allows some form nesting of iterations: during
an iteration over the input, the program can accumulate $\Diamond$s to
use for iteration over substructures that are smaller than the current
point in the iteration. A construction, due to \citet{AehligS02},
illustrates how this leads to all polytime computations. As above, we
assume a polynomial $p(n)$ and a step function
$f : \mathit{St} \multimap \mathit{St}$ that needs to be iterated
$p(n)$ times. We construct a linear iterator:
\begin{displaymath}
  \begin{array}{l}
    I_1 : (\tyNat \otimes \mathit{St}) \multimap (\tyNat \otimes \mathit{St}) \\
    I_1 = \lambda (n, s).\,\tmRec\,n\,\{
    \begin{array}[t]{@{}lcl}
      \conZero(d)&\mapsto&(\conZero(d), s);\\
      \conSucc(d,(n,s)) &\mapsto& (\conSucc(d,n),f\,s) \}
    \end{array}
  \end{array}
\end{displaymath}
Note that this iterator returns the natural number input as well as
the new state. LFPL does not allow duplication of iterable inputs, so
we must always reconstruct it if we want to do further
iteration. Addition of polynomials is accomplished by composition of
iterators. To raise the degree, we again use a nesting iterator:
\begin{displaymath}
  \begin{array}{l}
    I_{k+1} : (\tyNat \otimes \mathit{St}) \multimap (\tyNat \otimes \mathit{St}) \\
    I_{k+1} = \lambda (n,s).\,\tmRec\,n\,\{
    \begin{array}[t]{@{}lcl}
      \conZero(d)&\mapsto&(\conZero(d),s); \\
      \conSucc(d,(n,s)) &\mapsto& \Let\,(n,s)=I_k\,(n,s)\,\In\,(\conSucc(d,n), f\,s) \}
    \end{array}
  \end{array}
\end{displaymath}
Unlike in the \emph{Cons-free} system, this iterator does not raise
the degree of the nested iterator directly. Rather the iterator $I_k$
performs $\binom{n}{k}$ iterations. As observed by Aelhig and
Schwictenberg, this is sufficient because the binomials form a basis
for the vector space of all polynomials.

@@

%%
%% The acknowledgments section is defined using the "acks" environment
%% (and NOT an unnumbered section). This ensures the proper
%% identification of the section in the article metadata, and the
%% consistent spelling of the heading.
\begin{acks}
  Thanks to Anton Lorenzen and Fredrik Nordvall Forsberg for their
  comments on an earlier revision, and to the anonymous POPL reviewers
  for their detailed and interesting reviews. This work was funded by
  the \grantsponsor{EPSRC}{Engineering and Physical Sciences Research
    Council}{https://www.ukri.org/about-us/epsrc/}. Grant number
  \grantnum{EPSRC}{EP/T026960/1}, \emph{AISEC: AI Secure and
    Explainable by Construction}.
\end{acks}

>>>>

We will see in \autoref{sec:consfree-sound} that this system is sound
for polytime by a realisability argument. Completeness can be seen
more directly by constructing a function that iterates a function for
a statically known polynomial number of times in the size of the
input. Assume that we have a known polynomial
$p(n) = c_dn^d + \dots + c_0$ of degree $d$ with natural number
coefficients and some single step function
$f : \mathit{St} \multimap \mathit{St}$ over a state type
$\mathit{St}$ that runs to completion for input of size $n$ in $p(n)$
steps. Then, using the iterator above we can iterate $f$ over a
$\tyNat$ representing the size of the input:
\begin{displaymath}
  \begin{array}{l}
    I_1 : \tyNat \multimap \mathit{St} \multimap \mathit{St} \\
    I_1 = \lambda n. \lambda s.\tmRec\,n\,\{\conZero \mapsto s; \conSucc(s) \mapsto f\,s \}
  \end{array}
\end{displaymath}
To achieve higher degrees, we can use $\dupNat$ to nest iterations:
\begin{displaymath}
  \begin{array}{l}
    I_{k+1} : \tyNat \multimap \mathit{St} \multimap \mathit{St} \\
    I_{k+1} = \lambda n. \lambda s.
    \begin{array}[t]{@{}l}
      \Let\,(n,n') = \dupNat\,n\,\In\\
      \tmRec\,n\,\{\conZero \mapsto s; \conSucc(s) \mapsto I_k\,n'\,s \}
    \end{array}
  \end{array}
\end{displaymath}
By further use of $\dupNat$ and composition to handle addition of
polynomials, the function $f$ can now be iterated $p(n)$ many times,
where $n$ is the input $\tyNat$. Thus, the \emph{Cons-free} system can
represent all polytime computations.

@@

  \inferrule*
  {\Gamma_1 \vdash M : \Diamond \\ \Gamma_2 \vdash N : \tyNat}
  {\Gamma_1, \Gamma_2 \vdash \conSucc(M, N) : \tyNat}
\end{mathpar}
To construct a $\conZero$, we must have a $\Diamond$ to pay for it,
and likewise, to construct a $\conSucc$ we must pay a $\Diamond$. We
can think of $\Diamond$s as an unit of iterable data. Iterability is
``saved up'' in data during construction, and released during
iteration. Diamonds cannot be created by a program itself, for the
same reason that constructors were prohibited in the \emph{Cons-free}
system, but they are released from iterable data during iteration. The
LFPL natural number iterator has the following typing rule:
\begin{displaymath}
  \inferrule*
  {d : \Diamond \vdash M_z : A \\ d : \Diamond, x : A \vdash M_s : A \\ \Gamma \vdash N : \tyNat}
  {\Gamma \vdash \tmRec\,N\,\{\conZero(d) \mapsto M_z; \conSucc(d,x) \mapsto M_s \} : A}
\end{displaymath}
The difference with the \emph{Cons-free} iterator above is that the
$\conZero$ and $\conSucc$ cases now both have an additional binding of
type $\Diamond$. This allows some form nesting of iterations: during
an iteration over the input, the program can accumulate $\Diamond$s to
use for iteration over substructures that are smaller than the current
point in the iteration. A construction, due to \citet{AehligS02},
illustrates how this leads to all polytime computations. As above, we
assume a polynomial $p(n)$ and a step function
$f : \mathit{St} \multimap \mathit{St}$ that needs to be iterated
$p(n)$ times. We construct a linear iterator:
\begin{displaymath}
  \begin{array}{l}
    I_1 : (\tyNat \otimes \mathit{St}) \multimap (\tyNat \otimes \mathit{St}) \\
    I_1 = \lambda (n, s).\,\tmRec\,n\,\{
    \begin{array}[t]{@{}lcl}
      \conZero(d)&\mapsto&(\conZero(d), s);\\
      \conSucc(d,(n,s)) &\mapsto& (\conSucc(d,n),f\,s) \}
    \end{array}
  \end{array}
\end{displaymath}
Note that this iterator returns the natural number input as well as
the new state. LFPL does not allow duplication of iterable inputs, so
we must always reconstruct it if we want to do further
iteration. Addition of polynomials is accomplished by composition of
iterators. To raise the degree, we again use a nesting iterator:
\begin{displaymath}
  \begin{array}{l}
    I_{k+1} : (\tyNat \otimes \mathit{St}) \multimap (\tyNat \otimes \mathit{St}) \\
    I_{k+1} = \lambda (n,s).\,\tmRec\,n\,\{
    \begin{array}[t]{@{}lcl}
      \conZero(d)&\mapsto&(\conZero(d),s); \\
      \conSucc(d,(n,s)) &\mapsto& \Let\,(n,s)=I_k\,(n,s)\,\In\,(\conSucc(d,n), f\,s) \}
    \end{array}
  \end{array}
\end{displaymath}
Unlike in the \emph{Cons-free} system, this iterator does not raise
the degree of the nested iterator directly. Rather, the iterator $I_k$
on the input $n$ performs $\binom{n}{k}$ iterations. As observed by
Aelhig and Schwictenberg, this is sufficient because the binomials
form a basis for the vector space of all polynomials.

@@

%%
%% The acknowledgments section is defined using the "acks" environment
%% (and NOT an unnumbered section). This ensures the proper
%% identification of the section in the article metadata, and the
%% consistent spelling of the heading.
\begin{acks}
  Thanks to Anton Lorenzen and Fredrik Nordvall Forsberg for their
  comments on an earlier revision, and to the anonymous POPL reviewers
  for their detailed, perceptive, and interesting reviews. This work
  was funded by the \grantsponsor{EPSRC}{Engineering and Physical
    Sciences Research
    Council}{https://www.ukri.org/about-us/epsrc/}: Grant number
  \grantnum{EPSRC}{EP/T026960/1}, \emph{AISEC: AI Secure and
    Explainable by Construction}.
\end{acks}
